\large How resource abundance and stochasticity affect organisms' space-use requirements

### 2024-04-12 writing/manuscript.Rmd
addressed mike's comments; see ***changes***

---
title: \large How resource abundance and stochasticity affect organisms' space-use requirements
# author:
#   - name: Stefano Mezzini
#     institute: [braes, biol]
#   - name: Chris H. Fleming
#     institute: [ucf, smith]
#   - name: E. Patrícia Medici
#     institute: [tapir, iucn, escas]
#   - name: Michael J. Noonan
#     email: michael.noonan@ubc.ca
#     institute: [braes, biol, cmps]
#     correspondence: "yes"
# institute:
#   - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
#   - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
#   - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
#   - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
#   - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
#   - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
#   - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
#   - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'hr-resource-stoch.bib'
csl: 'the-american-naturalist.csl' # use 'nature.csl' to count refs
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{wrapfig} # for wrapping text around figures
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font={scriptsize, doublespacing}, labelfont=bf, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage[left]{lineno} # for line numbers
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # @reed_frequency_2003 has character U+0308
    pandoc_args:
# https://github.com/pandoc/lua-filters/blob/master/author-info-blocks/README.md
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
editor_options: 
  chunk_output_type: console
---

@@

\noindent \textbf{References}: 95 <!-- counted by using nature csl -->

@@

<!-- start main body on a new page -->
# To do:

* make sure appendices flow (added stuff from main text)
* check number of references (updated on 2024-03-18)

@@

\noindent where $H$ is the organism's home-range size, $C$ is the organism's resource consumption (kcal day$^{-1}$), and $R$ is the resources the organism can access (kcal day$^{-1}$ unit area$^{-1}$). Harestad and Bunnel's model is simple to conceptualize and allows for testable predictions, but it fails to account for the nonlinear relationship required to avoid negative home range estimates [e.g., fig. 2 in the work of @bista_effect_2022].<!-- to an organism's values of $C$ and $R$ likely depend on numerous factors, such as competition, metabolic rate, diet, and body weight [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982; @jetz_scaling_2004; @boratynski_energetic_2020; @noonan_effects_2020].--> Many researchers have since demonstrated that organisms adapt their home ranges in response to resources abundance, but studies often fail to build upon previous work. Instead, results are typically reported as independent, novel findings. Thus, there remains a need for a clear hypothesis for the effects of resource abundance on organisms' space-use requirements. Additionally, while much work has been done on estimating organisms' responses to average resource abundance, there is little information on how they respond to unpredictable changes in resources [but see: @stephens_optimal_1982; @rizzuto_forage_2021].

Here, we refer to a location's average amount of resources as "resource abundance", while we use the phrase "resource stochasticity" to indicate the variability in resources after accounting for changes in the mean. We argue that, on its own, a habitat's long-term resource abundance is not sufficient to assess the habitat's quality, nor make predictions about how much space an organism might use. For instance, herbivores in a grassland with relatively low but constant forage availability (low mean and variance) will adopt different behaviors and adaptations from herbivores in a desert with equally scarce forage but rare, sudden, and strong pulses of resources (low mean and high stochasticity). In the grassland, an animal may require a large but constant home range as it moves between locations in search of food [@teitelbaum_beyond_2019], while an animal in the desert may switch between dispersal in search for high-resource patches and short-term range residency whinith patches [@charnov_optimal_1976]. Although it may be possible for both habitats to have the same long-term average resource abundance, the differences in resource unpredictability result in different selection of movement and life history strategies. Although resource unpredictability (and seasonality) may ecrease organisms' fitness and a landscape's energetic balance [@chevin_adaptation_2010], there is still limited empirical evidence to support this hypothesis [but see: @herfindal_prey_2005; @nilsen_can_2005; @rizzuto_forage_2021].

Here, we illustrate how organisms' space use depends on both the abundance and unpredictability of resources. First, we set the theoretical background necessary for the successive sections by introducing key concepts and notation. Next, we provide a review of the effects of resource abundance on organisms' space use while suggesting a unifying hypothesis. Afterwards, we provide a review of the effects of resource stochasticity on organisms' space use while suggesting a second unifying hypothesis. Subsequently, we demonstrate the power of these two hypotheses using quantitative, simulated responses to changes in resource abundance and unpredictability. Finally, we demonstrate how this framework can be used in practice to describe the movement ecology of a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado [@medici_movement_2022]. Results from the simulations and empirical example are presented using a fully transparent approach that allows researchers to replicate the work and apply the methods to their own tracking data.

@@

\noindent Resources are often unpredictable (and difficult to quantify), since they depend on various factors which cannot be accounted for easily, including climate [@lindstedt_seasonality_1985; @schmidt_interplay_2020; @morellet_seasonality_2013], weather [@fjelldal_nightly_2021; @morellet_seasonality_2013], competitive pressure [@torrez-herrera_monkeys_2020; @rich_anthropogenic_2012], and differences in energetics at the individual [@schmidt_interplay_2020] and species level [@jetz_scaling_2004]. Thus, let the random variable $R$ indicate the amount of resources at a given point in space and time. Quantifying resources as a numerical random variable, as opposed to using *ad hoc* qualitative descriptions, provides us with the capacity to leverage techniques from probability theory and statistics. We assume readers are familiar with fundamental statistical concepts (e.g., random variables, probability distributions, expected value and variance of a random variable), but we define and explain the necessary concepts in Appendix A.

@@

\noindent While organisms' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction. All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@harvey_primate_1981; @baldwin_nutritional_1984; @reich_body_2001]. Failing to acquire sufficient resources can result in lower fitness, behavioral changes, physiological changes or damage, inability to reproduce, and death. Motile organisms can move to new locations in search of resources [@charnov_optimal_1976; @brown_ecology_1999], but they must also account for a higher metabolism and movement costs [@taylor_energetics_1982]. The hypothesis that space-use requirements decrease with resource abundance, $\e(R)$, is commonly accepted and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022; @bradsworth_using_2022; @mcclintic_effects_2014]. This is problematic because a linear relationship can lead to questionable estimates and problematic biases, such as negative or excessively small home ranges [e.g., fig. 2 in the work of @bista_effect_2022]. Conceptually, the relationship between space-use requirements and $\e(R)$ must be nonlinear, since there are limits to how much space an organism is able to explore in its lifetime and a minimum amount of space it requires to survive [which is necessarily greater than zero -- see: @lucherini_habitat_1996; @herfindal_prey_2005; @nilsen_can_2005; @simcharoen_female_2014; @watson_ferruginous_2020, and contrast them to the estimates based on linear models listed above].

@@

Fig. \@ref(fig:mu-hyp) illustrates our first of two hypotheses: When $\e(R)$ is high, we expect organisms' spatial requirements to be relatively small and near the smallest amount of space required to survive [e.g., @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005]. At lower values of $\e(R)$, we expect organisms' spatial requirements to increase nonlinearly, since low values of $\e(R)$ force organisms to expand their home ranges [@lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019]. It is unclear when organisms switch from range residency to migration or nomadism (or vice-versa), but understanding the connection between these types of movement is important for quantifying the effect of resource abundance on organisms' space-use requirements and when an organism may choose to migrate or disperse rather than remaining range-resident [mammals: @teitelbaum_how_2015; moose, *Alces alces*: @singh_migration_2012; eagles, *Haliaeetus leucocephalus*: @wheat_migrate_2017; @poessel_interpreting_2022; lesser flamingos, *Phoeniconaias minor*: @pretorius_movement_2020]. Still, species-level changes in movement behavior are more likely to occur over evolutionary timescales than over an organism's lifespan, since larger ranges require greater vagility, which, in turn, is facilitated by morphological features such as hinged joints and elongated limbs [@jetz_scaling_2004; @hirt_general_2017; @andersson_elbow-joint_2004; @samuels_postcranial_2013].

Consider the scenario where $\e(R) = \mu(t)$ changes over time but $\var(R) = \sigma^2 > 0$ is constant over time and space (fig. A1).<!-- We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant.--> Since $R$ is random, an organism that moves in the landscape will not find the same $R$ at different time points or throughout the habitat. However, if $\mu(t)$ changes repetitively and regularly over time (e.g. peaks each spring and is lowest in winter), an organism may learn to predict times of high or low $\e(R)$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019; @samarra_movements_2017] but will still not be able to predict $R$ exactly.

@@

\noindent The example illustrated in the section above can occasionally be a useful simplification of relatively stable environments or when data is scarce, but it likely not realistic, since $\var(R)$ often also changes over time. Periods of scarcity (e.g., winter, droughts) not only have lower $\e(R)$ but also have narrower ranges of $R$. In contrast, periods of high $\e(R)$ (e.g., spring green-up) also often tend to have higher $\var(R)$ because $R$ tends to vary across wider ranges (e.g., due to high spatial variability, or high $R$ followed by low $R$ after grazing). When both $\e(R)$ and $\var(R)$ change over time (fig. A2), predicting $R$ becomes more complex, since disentangling changes in $\e(R)$ and $\var(R)$ is not simple [@steixner-kumar_strategies_2020]. Statistically, this is because the more change one attributes to $\mu(t)$ (i.e., the wigglier it is), the smaller $\sigma^2(t)$ becomes. Conversely, the smoother $\mu(t)$ is, the larger $\sigma^2(t)$ becomes. Biologically, this is important because an organism's perception scale [@levin_problem_1992] determines whether it sees a change in $R$ as a trend in $\e(R)$ or as a stochastic event (i.e., due to $\var(R)$). An organism's perception of changes in $R$ will also depend strongly on the its cognitive capacities and memory [@mueller_social_2013; @abrahms_memory_2019; @foley_severe_2008; @falcon-cortes_hierarchical_2021; @fagan_spatial_2013]. Whether, an organism is able to predict trends in $\sigma^2(t)$ or not, environmental variability is thought to reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases organisms' fitness [e.g., @berger_climate_2018] and increases their space-use requirements. While this is true for both predictable and unpredictable stochasticity, extreme and rare events are more likely to have a stronger effect due to their unpredictability and magnitude [@logares_black_2012; @anderson_black-swan_2017]. A few recent studies support these hypotheses [@chevin_adaptation_2010; @morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], but many of them are limited in geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is still limited. Thus, there remains a need for developing a more complete understanding of how organisms' space-use requirements change with environmental stochasticity.

@@

\noindent We have provided the case for why both $\e(R)$ and $\var(R)$ affect organisms' space-use requirements, but we presented the two parameters as independent drivers of movement. However, organisms may respond to changes in $\var(R)$ more when resources are scarce than when they are abundant. Consequently, models estimating the effects of $R$ on organism's movement should account for not only the marginal effects of $\e(R)$ and $\var(R)$ but also their interactive effects. A highly unpredictable habitat may be very inhospitable if resources are poor, but $\var(R)$ may have little effect if resources are stochastic but always abundant. Thus, we expect $\var(R)$ to have a stronger effect on space-use requirements when $\e(R)$ is low, and less of an effect when $\e(R)$ is high. We explore this more in the following section.

@@

\noindent To support our hypotheses of how organisms' space use is affected by $\e(R)$, $\var(R)$, and the interaction effect of $\e(R)$ and $\var(R)$, we present the results from a series of quantitative simulations. To start, we used the `ctmm` package [@fleming_ctmm_2021] for `R` [@r_core_team_r_2023] to generate 200 tracks (see Appendix B for sensitivity analyses) from an Integrated Ornstein-Uhlenbeck movement model [IOU model, see @gurarie_correlated_2017]. The IOU model's correlated velocity produces realistic tracks with directional persistence despite the tracks being discrete samples of continuous-time movement processes. However, unlike Ornstein-Uhlenbeck (OU) and Ornstein-Uhlenbeck Foraging (OUF) models, IOU models do not produce spatially stationary movement, so the organism is not to range-resident. Consequently, each track is spatially unrestricted and can be interpreted as purely exploratory movement.

Each of the 200 tracks were placed on a raster with common starting point $\langle 0, 0\rangle$ and sufficient time between tracks to be independent of one another (other than the starting point). Each time the track moved to a new cell, the organism collected resources $R$ sampled from a Gamma distribution. The mean and variance of the distribution were defined by deterministic functions $\mu(t)$ and $\sigma^2(t)$ (orange and blue lines in fig. \@ref(fig:5-5-sims)). The value of $t$ was constant within each set of 200 tracks, so the time spent moving by the organism in each track did not affect the distribution $R$ was sampled from. Tracks were truncated once the organism reached satiety, and the organism was given enough time to return to $\langle 0, 0\rangle$ independently from the following track. Finally, as each set of 200 truncated was now spatially restricted and centered around $\langle 0, 0\rangle$, the set of tracks was modeled to calculate the 95% utilization distribution from an OUF model and Autocorrelated Kernel Density Estimate. We designed the simulations presented here to estimate the effects of $\e(R)$ and $\var(R)$ in simplistic environments. Consequently, we made the following four assumptions: (1) $\e(R)$ and $\var(R)$ are constant over space and short periods of time (i.e., within each set of 200 tracks, but not over $t$). (2) Organisms move randomly in the environment with no energetic costs. (3) Organisms have no perceptive range or memory, so they are unable to detect, learn, or predict where resources are abundant (high $\e(R)$) or reliable (low $\var(R)$) over time. (4) Resources do not deplete, and there is no competition or predator avoidance. Additional information is provided in Appendix B.

@@

Fig. \@ref(fig:5-5-sims) shows how simulated space-use requirements responded to changes in $\mu(t)$ and $\sigma^2(t)$ in scenarios where both functions can remain constant, increase linearly, oscillate cyclically, drift stochastically, or change erratically. The top row <!--(constant $\var(R)$) -->shows how space-use requirements vary for different trends in $\mu(t)$ while $\var(R)$ remains constant (like in fig. A1). As $\e(R)$ increases at a constant slope (linear $\mu(t)$) $H$ decreases nonlinearly, with larger changes when $\e(R)$ is low, until it approaches the minimum size required by the organism. Although less evident, note how the noise in the lines also decreases as $\e(R)$ increases.<!-- As described in the section on the effects of resource abundance, the nonlinear decrease is because changes in $\mu(t)$ have a larger effect when $\e(R)$ is low [e.g., @lucherini_habitat_1996; @nilsen_can_2005; @herfindal_prey_2005].-->

@@

\noindent This paper presents two hypotheses for describing the effects of resource abundance and stochasticity on organisms' space-use requirements. We support the hypotheses using quantitative simulations and an empirical case study. Each of the two methods allows us to highlight the strengths and limitations of simulation-based and data-driven approaches, which we discuss below.

>>>>

---
title: \large How resource abundance and stochasticity affect organisms' space-use requirements
# author:
#   - name: Stefano Mezzini
#     institute: [braes, biol]
#   - name: Chris H. Fleming
#     institute: [ucf, smith]
#   - name: E. Patrícia Medici
#     institute: [tapir, iucn, escas]
#   - name: Michael J. Noonan
#     email: michael.noonan@ubc.ca
#     institute: [braes, biol, cmps]
#     correspondence: "yes"
# institute:
#   - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
#   - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
#   - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
#   - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
#   - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
#   - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
#   - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
#   - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'hr-resource-stoch.bib'
csl: 'the-american-naturalist.csl'
# csl: 'nature.csl' # to count refs
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{wrapfig} # for wrapping text around figures
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font={scriptsize, doublespacing}, labelfont=bf, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage[left]{lineno} # for line numbers
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # @reed_frequency_2003 has character U+0308
    pandoc_args:
# https://github.com/pandoc/lua-filters/blob/master/author-info-blocks/README.md
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
editor_options: 
  chunk_output_type: console
---

@@

\noindent \textbf{References}: 102 (updated on 2024-04-12) <!-- counted by using nature csl -->

@@

<!-- start main body on a new page -->

@@

\noindent where $H$ is the organism's home-range size, $C$ is the organism's resource consumption (kcal day$^{-1}$), and $R$ is the resources the organism can access (kcal day$^{-1}$ unit area$^{-1}$). Harestad and Bunnel's model is simple to conceptualize and allows for testable predictions, but it fails to account for the nonlinear relationship required to avoid negative home range estimates [e.g., fig. 2 in the work of @bista_effect_2022].<!-- to an organism's values of $C$ and $R$ likely depend on numerous factors, such as competition, metabolic rate, diet, and body weight [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982; @jetz_scaling_2004; @boratynski_energetic_2020; @noonan_effects_2020].--> Many researchers have since demonstrated that organisms adapt their home ranges in response to resources abundance, ***but, unlike other areas of ecological research, findings from these studies are not structured around a set of theoretical expectations.*** Instead, results are typically reported as independent, novel findings. Thus, there remains a need for a clear hypothesis for the effects of resource abundance on organisms' space-use requirements. Additionally, while much work has been done on estimating organisms' responses to average resource abundance, there is little information on how they respond to unpredictable changes in resources [but see: @stephens_optimal_1982; @rizzuto_forage_2021].

Here, we refer to a location's average amount of resources as "resource abundance", while we use the phrase "resource stochasticity" to indicate the variability in resources after accounting for changes in the mean. We argue that, on its own, a habitat's long-term resource abundance is not sufficient to assess the habitat's quality, nor make predictions about how much space an organism might use. For instance, herbivores in a grassland with relatively low but constant forage availability (low mean and variance) will adopt different behaviors and adaptations from herbivores in a desert with equally scarce forage but rare, sudden, and strong pulses of resources (low mean and high stochasticity). In the grassland, an animal may require a large but constant home range as it moves between patches in search of food [@teitelbaum_beyond_2019], while an animal in the desert may switch between dispersal in search for high-resource patches and short-term range residency ***within*** patches [@charnov_optimal_1976]. Although the two habitats may have the same long-term average resource abundance, the ephemerality of resource in the second environment would promote a different selection of movement and life history strategies. Overall, resource stochasticity may decrease organisms' fitness and landscapes' energetic balances [@chevin_adaptation_2010], but there is still limited empirical evidence to support this hypothesis [but see: @herfindal_prey_2005; @nilsen_can_2005; @rizzuto_forage_2021].

Here, we illustrate how organisms' space use ***should be expected to*** depend on both the abundance and unpredictability of resources. First, we set the theoretical background necessary for the successive sections by introducing key concepts and notation. Next, we provide a review of the effects of resource abundance on organisms' space use while suggesting a unifying hypothesis. Afterwards, we provide a review of the effects of resource stochasticity on organisms' space use while suggesting a second unifying hypothesis. Subsequently, we demonstrate the power of these two hypotheses using quantitative, simulated responses to changes in resource abundance and unpredictability. Finally, we demonstrate how this framework can be used in practice to describe the movement ecology of a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado [@medici_movement_2022]. Results from the simulations and empirical example are presented using a fully transparent approach that allows researchers to replicate the work and apply the methods to their own tracking data.

@@

\noindent Resources are often unpredictable (and difficult to quantify), since they depend on various factors which cannot be accounted for easily, including climate [@lindstedt_seasonality_1985; @schmidt_interplay_2020; @morellet_seasonality_2013], weather [@fjelldal_nightly_2021; @morellet_seasonality_2013], competitive pressure [@torrez-herrera_monkeys_2020; @rich_anthropogenic_2012], and differences in energetics at the individual [@schmidt_interplay_2020] and species level [@jetz_scaling_2004]. ***Thus, we can treat the amount of resources at a given point in space ($u$) and time ($t$) as a random variable, denoted $R(u, t)$. The benefit of treating*** resources as a numerical random variable, ***is that it*** provides us with the capacity to leverage techniques from probability theory and statistics. We assume readers are familiar with fundamental statistical concepts (e.g., random variables, probability distributions, expected value and variance of a random variable), but we define and explain the necessary concepts in Appendix A.

@@

\noindent While organisms' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction. All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@harvey_primate_1981; @baldwin_nutritional_1984; @reich_body_2001]. ***As the abundance of essential resources fluctuates,*** motile organisms can move to new locations ***or 'patches', to meet their requirements*** [@charnov_optimal_1976; @brown_ecology_1999], but they must also account for a higher metabolism and movement costs [@taylor_energetics_1982].

@@

Fig. \@ref(fig:mu-hyp) illustrates our first of two hypotheses: When $\e(R)$ is high, we expect organisms' spatial requirements to be relatively small and near the smallest amount of space required to survive [e.g., @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005]. At lower values of $\e(R)$, we expect organisms' spatial requirements to increase nonlinearly, since low values of $\e(R)$ force organisms to expand their home ranges [@lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019]. It is unclear when organisms switch from range residency to migration or nomadism (or vice-versa), but understanding the connection between these types of movement is important for quantifying the effect of resource abundance on organisms' space-use requirements and when an organism may choose to migrate or disperse rather than remaining range-resident [mammals: @teitelbaum_how_2015; moose, *Alces alces*: @singh_migration_2012; eagles, *Haliaeetus leucocephalus*: @wheat_migrate_2017; @poessel_interpreting_2022; lesser flamingos, *Phoeniconaias minor*: @pretorius_movement_2020]. Still, species-level changes in movement behavior are more likely to occur over evolutionary timescales than over an organism's lifespan, since larger ranges require greater vagility, which, in turn, is facilitated by morphological features such as hinged joints and elongated limbs [@jetz_scaling_2004; @hirt_general_2017; @andersson_elbow-joint_2004; @samuels_postcranial_2013]. ***moved from above*** The hypothesis that space-use requirements decrease with resource abundance, $\e(R)$, is commonly accepted and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022; @bradsworth_using_2022; @mcclintic_effects_2014]. This is problematic because, ***conceptually,*** the relationship between space-use requirements and $\e(R)$ must be nonlinear, since: (1) there ***is an upper limit*** to how much space an organism is able to explore in its lifetime and ***(2)*** the minimum amount of space it requires to survive ***is necessarily greater than zero*** [see: @lucherini_habitat_1996; @herfindal_prey_2005; @nilsen_can_2005; @simcharoen_female_2014; @watson_ferruginous_2020, and contrast them to the estimates based on linear models listed above].

Consider the scenario where $\e(R) = \mu(t)$ changes over time but $\var(R) = \sigma^2 > 0$ is constant over time and space (fig. A1). Since $R$ is random, an organism that moves in the landscape will not find the same $R$ at different time points or throughout the habitat. However, if $\mu(t)$ changes repetitively and regularly over time (e.g. peaks each spring and is lowest in winter), an organism may learn to predict times of high or low $\e(R)$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019; @samarra_movements_2017] but will still not be able to predict $R$ exactly.

@@

\noindent The example illustrated in the section above can occasionally be a useful simplification of relatively stable environments or when ***information on how $\e(R)$ changes is limited,*** but it likely not realistic, since ***the variance in $R$ around $\e(R)$, which we indicate as $\var(R)$, can also differ between locations or time periods***. Generally, periods of scarcity (e.g., winter, droughts) not only have lower $\e(R)$ but also have narrower ranges of $R$***, which implies a smaller $\var(R)$***. In contrast, periods of high $\e(R)$ (e.g., spring green-up) also often tend to have higher $\var(R)$, ***as*** $R$ tends to vary ***more (e.g., due differences in phenology, environmental conditions, and patch depletion by competitors). Recognizing differences in $\var(R)$ between locations or periods helps account for the small-scale stochasticity in $R$, after accounting for the large-scale average $R$ [@levin_problem_1992]. However,*** when both $\e(R)$ and $\var(R)$ change over time (fig. A2), predicting $R$ becomes more complex, since disentangling changes in $\e(R)$ and $\var(R)$ is not simple [@steixner-kumar_strategies_2020]. Statistically, this is because the more change one attributes to $\mu(t)$ (i.e., the wigglier it is), the smaller $\sigma^2(t)$ becomes. Conversely, the smoother $\mu(t)$ is, the larger $\sigma^2(t)$ becomes. Biologically, this is important because an organism's perception scale [@levin_problem_1992] determines whether it sees a change in $R$ as a trend in $\e(R)$ or as a stochastic event (i.e., due to $\var(R)$). An organism's perception of changes in $R$ will also depend strongly on the its cognitive capacities and memory [@mueller_social_2013; @abrahms_memory_2019; @foley_severe_2008; @falcon-cortes_hierarchical_2021; @fagan_spatial_2013]. Whether, an organism is able to predict trends in $\sigma^2(t)$ or not, environmental variability is thought to reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases organisms' fitness [e.g., @berger_climate_2018] and increases their space-use requirements. While this is true for both predictable and unpredictable stochasticity, extreme and rare events are more likely to have a stronger effect due to their unpredictability and magnitude [@logares_black_2012; @anderson_black-swan_2017]. A few recent studies support these hypotheses [@chevin_adaptation_2010; @morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], but many of them are limited in geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is ***unknown***. Thus, there remains a need for developing a more complete understanding of how organisms' space-use requirements change with environmental stochasticity.

@@

\noindent We have provided the case for why both $\e(R)$ and $\var(R)$ ***should be expected to*** affect organisms' space-use requirements, but we presented the two parameters as independent drivers of movement. However, organisms may respond to changes in $\var(R)$ more when resources are scarce than when they are abundant. Consequently, ***an organism’s movement behavior is likely to be a function of*** not only the marginal effects of $\e(R)$ and $\var(R)$ but also their interactive effects. A highly unpredictable habitat may be very inhospitable if resources are poor, but $\var(R)$ may have little effect if resources are stochastic but always abundant. Thus, we expect $\var(R)$ to have a stronger effect on space-use requirements when $\e(R)$ is low, and less of an effect when $\e(R)$ is high. We explore this more in the following section.

@@

\noindent To support our hypotheses of how organisms' space use is affected by $\e(R)$, $\var(R)$, and the interaction effect of $\e(R)$ and $\var(R)$, we present the results from a series of quantitative simulations. To start, we used the `ctmm` package [@fleming_ctmm_2021] for `R` [@r_core_team_r_2023] to generate 200 tracks (see Appendix B for sensitivity analyses) from an Integrated Ornstein-Uhlenbeck movement model [IOU model, see @gurarie_correlated_2017]. The IOU model's correlated velocity produces realistic tracks with directional persistence, but, unlike Ornstein-Uhlenbeck (OU) and Ornstein-Uhlenbeck Foraging (OUF) models, IOU models do not produce spatially stationary movement, so the organism is not to range-resident. Consequently, each track is spatially unrestricted and can be interpreted as purely exploratory movement.

Each of the 200 tracks were placed on a raster ***(this was a raster. is ``grid" more understandable?)*** with common starting point $\langle 0, 0\rangle$ and sufficient time between tracks to be independent of one another (other than the starting point). Each time the track moved to a new cell, the organism collected resources $R$ sampled from a Gamma distribution. The mean and variance of the distribution were defined by deterministic functions $\mu(t)$ and $\sigma^2(t)$ (orange and blue lines in fig. \@ref(fig:5-5-sims)). The value of $t$ was constant within each set of 200 tracks, so the time spent moving by the organism in each track did not affect the distribution $R$ was sampled from. Tracks were truncated once the organism reached satiety, and the organism was given enough time to return to $\langle 0, 0\rangle$ independently from the following track. Finally, as each set of 200 truncated was now spatially restricted and centered around $\langle 0, 0\rangle$, the set of tracks was modeled to calculate the 95% utilization distribution from an OUF model and Autocorrelated Kernel Density Estimate. We designed the simulations presented here to estimate the effects of $\e(R)$ and $\var(R)$ in simplistic environments. Consequently, we made the following ***five*** assumptions: (1) $\e(R)$ and $\var(R)$ are constant over space and short periods of time (i.e., within each set of 200 tracks, but not over $t$). (2) Organisms move randomly in the environment with no energetic costs. (3) Organisms have no perceptive range or memory, so they are unable to detect, learn, or predict where resources are abundant (high $\e(R)$) or reliable (low $\var(R)$) over time. (4) Resources do not deplete, and there is no competition or predator avoidance. ***(5) Since $\mu(t)$ and $\sigma^2(t)$ are spatially homogeneous for a given $t$, organisms respond to $\e(R)$ and $\var(R)$ by adapting the extent of the spatial needs (i.e., their positional variance), rather than dispersing or abandoning a range (i.e., their mean location). Additional information is provided in Appendix B, including the directed acyclical graph [see figure B6 and @mcelreath_statistical_2016] we used to infer causal the mechanisms of changes in $H$ and estimate the direct, partial effects of $\e(R)$ and $\var(R)$ on $H$ (contrast this with figure C5 and the empirical case study below).***

@@

Fig. \@ref(fig:5-5-sims) shows how simulated space-use requirements responded to changes in $\mu(t)$ and $\sigma^2(t)$ in scenarios where both functions can remain constant, increase linearly, oscillate cyclically, drift stochastically, or change erratically. The top row <!--(constant $\var(R)$) -->shows how space-use requirements vary for different trends in $\mu(t)$ while $\var(R)$ remains constant (like in fig. A1). As $\e(R)$ increases at a constant slope (linear $\mu(t)$) $H$ decreases nonlinearly, with larger changes when $\e(R)$ is low, until it approaches the minimum size required by the organism. Although less evident, note how the noise in the ***green*** lines also decreases as $\e(R)$ increases.<!-- As described in the section on the effects of resource abundance, the nonlinear decrease is because changes in $\mu(t)$ have a larger effect when $\e(R)$ is low [e.g., @lucherini_habitat_1996; @nilsen_can_2005; @herfindal_prey_2005].-->

@@

\noindent This paper presents two hypotheses ***that describe*** the effects of resource abundance and stochasticity on organisms' space-use requirements. ***We show that, together, these hypotheses provide a framework for understanding how organisms adapt their spatial needs to both average conditions and the variance around the average conditions. By moving beyond considering only average conditions, we are able to create a more realistic view of how animals respond to changes in both predictable and unpredictable environmental conditions. We support the hypotheses using quantitative simulations and test it using a simple case study with methods that can implemented easily by animal movement ecologists (see Appendix C). In this section, we discuss the strengths and limitations of both the simulation-based and empirically driven approaches along with comments on the consequences for free-moving animals in a changing world.***
